---
title: "CNTNAP_Marble_Analysis"
author: "Alexandra Phillips"
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_float: yes
    code_folding: hide
  pdf_document:
    toc: yes
---
# Overview 


The marble burying test measures anxiety and obsessive-compulsive behavior as under stress the mice will bury more marbles in the bedding. 



Load libraries 
```{r message=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(ggpubr)
library(ggplot2)
require(reshape2)
library(cowplot)
library(Hmisc)
library(mltools)
library(plyr)

```

Load dataset and pre-work
```{r echo=FALSE}

marble <- read.csv(file = "../data/behavior/Marbles_07052021.csv")
#marble <- subset(marble, Buried != "NA")

#Setting up some color 
sgColorPalette <- c("#84CF04","#01B5BB","#E50E63","#6D7272","#8F389E","#DF8236","#036B6B","#F1BA2F","#9F832D","#94E804" ,"#01D4DB","#FAC131","#B0B8B8","#F08C3A","#FF106E","#B948CC","#05B5B5","#CFAA3A")

marble$Compound <- as.factor(marble$Compound)
marble$Compound <- relevel(marble$Compound, "Control")

#exclude mice from training cohort 
marble <- marble[-c(40,41, 42, 43, 44, 45, 46, 47), ]
#not the best way to do it, but it works. 

my_comparisons <- list( c("5D", "Control"), c("Control", "UDC"), c("Control", "GDC"), c("Control", "CIT"), c("Control", "GLUT"), c("Control", "BL6"))
```


# Boxplot of buried marbles 
```{r echo=FALSE}
buried <- ggplot(marble, aes(x=Compound, y=Buried)) + 
   geom_boxplot(aes(fill=Compound), width = 0.5)+
  scale_fill_manual(values=sgColorPalette)+
  geom_jitter(width = 0.25)+  
  stat_compare_means(comparisons = my_comparisons, method = "wilcox.test") +
  ggtitle("Buried marbles") +
  scale_x_discrete(labels=c("Control n=15", "5D n=12", "BL6 n=12", "CIT n=9", "GDC n= 12","GLUT n=8", "UDC n=12"))
buried

# ANOVA
interaction <- aov(Buried ~ Compound, data = marble)
summary(interaction)

tukey.test <- TukeyHSD(interaction)
tukey.test
```
## Conclusion 


There are no significant different between the controls and metabolites via ANOVA, however mice fed ursodeoxycholate seem to bury more marbles and have less variation. The control cohort continues to have large variation now with two control cohorts combined. 

\newline
Mice fed glutamate have a large variation in the number of marbles buried. Mice fed citrate look very similar to the controls. 

## Conclusion 


The variation is not cohort dependent and not significantly different. Both control cohorts have a large amount of spread. 

# Barplot of buried marbles 
```{r warning=FALSE, echo=FALSE}
# Barplot Buried Marbles VS Compound
ggplot(marble, aes(x= Compound, y= Buried, fill = Compound)) +
  stat_summary(geom = "bar", fun.y = mean, position = "dodge", width=0.6) +
  stat_summary(geom = "errorbar", fun.data = mean_se, position = "dodge", width=0.6)+
  scale_fill_manual(values=sgColorPalette)+
  ggtitle("Buried marbles per compound")

```
